---
title: "Statistical models for the BIRDIE pipeline"
author: "Francisco Cervantes Peralta and Res Altwegg"
date: "7/07/2021"
subtitle: 'JRS Biodiversity Series'
output: beamer_presentation
classoption: "aspectratio=169"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

::: columns

:::: column

- example of a two-column slide

::::

:::: column

\begin{center}
\includegraphics[width=\textwidth]{SEEC_03_B.pdf}
\end{center}

\vspace{0.5cm}

http://www.seec.uct.ac.za/

\vspace{0.5cm}

@SEEC_UCT

::::

:::

## Occupancy: the proportion of sites occupied by a species


```{r occupied, out.width="50%", include=TRUE, fig.align="center", echo=F, fig.cap = "Some figure legend..."}
knitr::include_graphics("occupied.pdf")
```

## Occupancy: the proportion of sites occupied by a species}

\begin{columns}[]
\column{0.5\textwidth}{
\includegraphics[height=5.5cm]{occupied.pdf}
}
\column{0.5\textwidth}{
\begin{itemize}
\item Occupancy: $\Psi = \frac{occupied}{total}$
\item $logit(\Psi) = f(covariates)$
\end{itemize}
}
\end{columns}


## The species is not detected in all occupied cells

Detection probability $p < 1$

\includegraphics[height=4cm]{detections.pdf}

`Naive approach':
\begin{itemize}
\item $\Psi \times p = \frac{occupied}{total}$
\item $logit(\Psi \times p) = f(covariates)$
\end{itemize}


## The species is not detected in all occupied cells
\begin{columns}[]
\column{0.6\textwidth}{
\includegraphics[height=6.5cm]{detections2.pdf}
}
\column{0.4\textwidth}{

Repeated sampling

\vspace{0.5cm}

{\bf Assumptions:}
\begin{itemize}
\item Closure (no colonisation or exctinction)
\item No false detections
\end{itemize}
}
\end{columns}


## The species is not detected in all occupied cells
\begin{columns}[]
\column{0.6\textwidth}{
\includegraphics[height=6.5cm]{detections2.pdf}
}
\column{0.4\textwidth}{

{\bf Survey histories:}

1 = detected \\
0 = not detected
\begin{itemize}
\item (1,1) 000
\item (1,3) 111
\item (2,2) 001
\item (1,6) 000
\end{itemize}
}
\end{columns}



## The species is not detected in all occupied cells}

::: columns

:::: column

\includegraphics[height=6.5cm]{detections2.pdf}

::::

:::: column

**Survey histories:**

1 = detected

0 = not detected

\begin{itemize}
\item (1,1) 000
\item (1,3) 111
\item (2,2) 001
\item {\bf (1,6) 000}
\end{itemize}

::::

:::

How many occupied cells have no detections?




## A model for the detections

- $\Psi =$ probability of a cell to be occupied
- $p =$ probability of detecting the species given that the cell is occupied
- $K =$ number of visits to each site
- $y_i =$ number of detections at site $i$


\vspace{1cm}

$$Pr(Y=y_i) = \Psi \binom{K}{y_i} p^{y_i} (1-p)^{K-y_i}, y_i > 0$$
$$= \Psi(1-p)^K + (1-\Psi), y_i = 0$$




## Literature


::: columns

:::: column

- one paper


::::

:::: column

- another paper

::::

:::




