---
title: "BIRDIE ABU: Model summary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE ABU: Model summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```

## Introduction

The species abundance module (ABU) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species abundance* vignettes for general details
about BIRDIE and about the ABU module, respectively. In this vignette, we will go through
the different tasks that are performed during the first step of the ABU module: **model summary**.

The main function used for summarising a model fit is `ppl_summarise_ssm()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work.

Model predictions for latent population states (abundance and rates of change) are prepared by JAGS during
model fitting. Therefore, the only task that needs to be performed here is preparing
these predictions as a dataframe and get them ready for plotting. As side products
we will produce `.png` images for each species and site that will help in model
assessment.

The function that performs these tasks is `plotJagsSsm2ss()`, which is found in
`R/utils-jags.R`. For more details on how this function works see `?plotJagsSsm2ss`.
The output of this function give us a dataframe with predictions on the estimated abundance and rate
of change of the summer and winter populations for each species. Each dataframe
contains all the suitable sites (see *BIRDIE ABU: data preparation*) the species was counted at.
We are currently predicting abundance in the linear scale (the scale of the data),
but there is also the option to produce these estimates in the log scale. This is a
better option to assess model fit.
