---
title: "BIRDIE-DST: Model summary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE-DST: Model summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```

## Introduction

The species distributions module (DST) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species distributions* vignettes for general details
about BIRDIE and about the DST module, respectively. In this vignette, we will go through
the different tasks that are performed during the step of the DST module: **model summary**.

The main function used for summarising a model fit is `ppl_summarise_occu()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work.

This step is relatively simple. Here we need to conduct two tasks: predict occurrence and
detection probabilities for all pentads in South Africa from a fitted model and summarise
these predictions.

## Predicting from a model fit

`predictSpOccu()` is used for predicting from a model fit and it is basically a wrapper
around `spOccupancy::predict.PGOcc()`. However, it has a few data preparation steps
to make sure the data we pass on to the predict function has the same variables and
on the same scale as the data we used for fitting the model. Model are fitted using
only those pentads that were visited in any given year, but we want to predict for all
pentads in South Africa. To maintain the same scale, we make use of the covariate
scale information that we stored in the model fit object (see *BIRDIE DST: model fitting* vignette).

The output of `predictSpOccu()` is a list with two elements containing posterior
predictive samples for $\psi$ (psi, probability of occurrence) and $p$ (probability of
detection) for each pentad.


## Summarising predictions

In this step, we use the function `summariseSpOccu()` to extract the 0.025, 0.5 and 0.975
quantiles of the predictive samples for $\psi$ and $p$ obtained from `predictSpOccu()`.
We will store these quantiles to display on the BIRDIE website rather than all of the
posterior predictive samples.

We also compute the realized occupancy from the posterior predictive samples and the
data. The realized occupancy is the probability of occurrence conditional on the
observed data, such that

$$P[occu | obs = 1] = 1$$
$$P[occu | obs = 0] = \frac{\psi q}{(1 - \psi + \psi q)}$$
where $obs = 1$ when the species was detected at a site on any visit, and $obs = 0$
when the species was not detected at a site on any visit, $\psi$ is the probability of
occurrence estimated by our model and $q = \prod_{i=1}^N (1 - p_i)$, with $p_i$ being
the probability of detection on visit $i$ and $N$ the total number of visits to 
a given site.

All these summaries are stored in a file called `analysis/output/occu_pred_spOccupancy_YYYY_sp_code.csv`,
where `YYYY` is the year of the data we are fitting a model to and `sp_code` is the
species code. These files will be exported to the database for storage and display
on the BIRDIE website.

If no model could be fitted because of lack of enough data, then, we
use the raw data to create the prediction file mentioned above
The file still has the exact same structure, but now $\psi$ and $p$ will be `NA` for all
pentads (because they could not be calculated). Raw data will be treated as realized occupancy and no confidence limits will be estimated. The function that performs these tasks is `createPredFromAbap()`.
