---
title: "BIRDIE DST: Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE DST: Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```


## Introduction

The species distributions module (DST) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species distributions* vignettes for general details
about BIRDIE and about the DST module, respectively. In this vignette, we will go through
the different tasks that are performed during the step of the DST module: **model diagnostics**.

The main function used for model diagnosis is `ppl_diagnose_occu()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work.

During the diagnostics step we look at two indicators: the Gelman-Rubin convergence
diagnostic ($\hat{R}$) and a goodness-of-fit Bayesian p-value. The first diagnostic is
computed by the function `diagnoseRhatSpOccu()` and the second one by the function
`diagnoseGofSpOccu()` (both functions are found in `R/utils-spOccupancy.R`).

`ppl_diagnose_occu()` combines the output of these two functions into a single table
that is stored in `analysis/output/sp_code/occu_diagnostics_spOccupancy_YYYY_sp_code.csv`,
where `YYYY` is the year of the data we are fitting a model to and `sp_code` is the
species code. In this way, it is easy to extract all diagnostic files from the different
directories, combine them into a table and quickly explore potential issues.

## Diagnosing convergence

The function `diagnoseRhatSpOccu()` extracts $\hat{R}$ values computed by `spOccupancy`
for each of the estimated parameters and tabulates them. In this table, each parameter
appears in one column and gets a 0 if $|1 - \hat{R}| < 0.1$ and a 1 otherwise. Some
extra columns are added, such as number of non-convergent parameters, number of detections
observed, number of sites visited, etc.

## Diagnosing goodness-of-fit

The function `diagnoseGofSpOccu()` conducts posterior predictive checks (PPC) and calculates
a Bayesian p-value. This function is a modification of the `spOccupancy::ppcOcc()`
making a slightly more efficient use of RAM memory, but at the expense of adding
some dependencies (on which BIRDIE already depended anyway, notably [`dplyr`](https://github.com/tidyverse/dplyr)).

The procedure to estimate the Bayesian p-value is the same used by `spOccupancy::ppcOcc()`.
However `diagnoseGofSpOccu()` only conducts chi-squared test at the moment, and it
only aggregates data per site and not per visit. With `spOccupancy::ppcOcc()` one
can aggregate data per site or visit and conduct Freeman-Tuckey and Chi-squared tests.

The procedure consists of simulating a number of detection/non-detection data sets
from the posterior distribution and comparing the number of simulated detections, with
the number of detections actually observed in the data (see Doser et al. 2022 and references
therein). Each visit results in either
detection or non-detection, so what we really compare is the number of detections
in each site (i.e., we aggregate detections by site).

To make the comparison we first calculate a Chi-squared statistic for the data, for each
Monte Carlo Markov Chain (MCMC) iteration $i$

$$\chi_{obs}^{2(i)} = \sum_{y=1}^N\frac{(\textrm{Observed}_y - \textrm{Expected}_y^{(i)})^2}{\textrm{Expected}_y^{(i)}}$$
where $\textrm{Observed}_y$ is the number of detections observed at site $y$ and 
$\textrm{Expected}_y^{(i)}$ is the expected number of detections estimated for site
$y$ during MCMC iteration $i$.

We do the same thing with the simulated data sets. For the data set simulated from 
MCMC iteration $i$ we compute

$$\chi_{sim}^{2(i)} = \sum_{y=1}^N\frac{(\textrm{sim}_y^{(i)} - \textrm{Expected}_y^{(i)})^2}{\textrm{Expected}_y^{(i)}}$$
A Bayesian p-value is computed as the proportion of $\chi_{sim}^{2} > \chi_{obs}^{2}$.


## Shiny apps

We have developed a Shiny app that allows us to extract diagnostics for any species
and year and explore them in an interactive way. The app is still experimental and
can be found at `analysis/tools/diagnose_dst`
