---
title: "BIRDIE DST: Model fitting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE DST: Model fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=FALSE}
library(BIRDIE)
```


## Introduction

The species distributions module (DST) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species distributions* vignettes for general details
about BIRDIE and about the DST module, respectively. In this vignette, we will go through
the different tasks that are performed during the step of the DST module: **model fitting**.

The main function used for model fitting is `ppl_fit_occu_model()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work. The output of
`ppl_fit_occu_model()` is always a model fit but it will change depending on the
package we have selected in the `configPipeline()` function. Currently, there are two
options [`spOccupancy`](https://github.com/doserjef/spOccupancy) (preferred) and
[`occuR`](https://github.com/r-glennie/occuR) (not maintained, so probably not working).

In the following section, we explain the process of fitting a model using `spOccupancy`.

## Fitting a model using the `spOccupancy` package

The main function used for fitting occupancy models using the `spOccupancy` package is
`fitSpOccu()`, which is found in the `R/utils-spOccupancy.R` file.

The `spOccupancy` package gives us the option to fit single season and multi-season
occupancy models. We are **currently only fitting single season models**, because
there are quite a lot of data in a single season and therefore multi-season models
would demand large amounts of computing resources. Although we could think of setting up analyses
on a high-performance computing facility, we would like to keep hardware requirements
at reasonable levels for any institution to be able to run the pipeline. In addition,
we plan to update the pipeline yearly and therefore it seems reasonable to run
single season models with the new data.

We also have the option to run models with or without spatial random effects. We are
**currently running models without spatial effects**. We are using around 10 environmental
covariates in our models and our diagnostics indicate that are not necessary. The
pipeline could accommodate spatial models and the code structure is there to implement
them. However, it would require some work to finalise the code.

The `fitSpOccu()` function with some helpers (also found in `R/utils-spOccupancy.R`)
performs several tasks:

1. Format the site and visit data to fit in the `spOccupancy` package. The function
`prepSpOccuData_single()` does this at the moment, because we are running single-season
models. There is an homologous `prepSpOccuData_multi()` for multi-season models, but
it probably needs some edits to work. These functions are basically wrappers around
[`ABAP::abapToSpOcc_single()`](https://github.com/AfricaBirdData/ABAP/blob/main/R/abapToSpOcc_single.R)
and [`ABAP::abapToSpOcc_multi()`](https://github.com/AfricaBirdData/ABAP/blob/main/R/abapToSpOcc_multi.R).

2. Define priors and initial values. For species we don't have any models fit, we use
the default priors from `spOccupancy` (see `?spOccupancy::PGOcc` for details). However,
if we have model fits for previous years, we center the priors in the previous parameter
estimate (if it exists - we might have fitted a different model - more of this later)
and use the standard deviation of the estimate plus one, unless it is larger than the 
`spOccupancy` default (~ sqrt(2.5)), in which case we use the latter. An important
function to this is `defineSpOccuPriors()` (in `R/utils-spOccupancy.R`).

3. Run the model. We fit the model using `spOccupancy::PGOcc()` since we are not currently
using spatial effects. If we did, then we would use `spOccupancy::spPGOcc()`. Finally,
we add information about priors and covariate scales to the model fit to be used 
in other operations later.


## Different models for different species and years

It might be the case that the first model we try does not converge or does not
fit the data well (see vignette about diagnostics). At the moment of writing we have
defined two models: one with random effects for observer,
and one with random effects for observers and sites on detection probabilities. The
idea is, we first try the simplest model and if there are issues we gradually increase complexity.
These two models work fine for now, but in the future we might want to implement others.

The procedure to run multiple models at the moment manual. We define the models we
want to try in the control script (see the *BIRDIE: species distributions* vignette),
we run the pipeline for all species in a given year with the first model, then we run diagnostics,
select problem species, then run the pipeline for those problem species with the second model.
This process can surely be automated, but the functionality still needs to be developed.
