---
title: "BIRDIE: Species abundance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE: Species abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```

## Introduction

There are two main analytical modules in BIRDIE: distribution and abundance.

In this document we will see how to use the abundance (ABU) module. To do this we will 
have a look at the control script used to run the module and then we will break down
its different sections to understand what all the functions involved in this analysis do.

## The control script

The control script can be found at `/analysis/scripts/pipeline_script_abu.R`. This script
has three parts:

- Configuration
- Create logs
- Run modules

We will go through each of these parts below.

```{r, echo=FALSE, eval=FALSE}
writeLines(readLines("../analysis/scripts/pipeline_script_abu.R"))
```

The script is really just a for loop over species in which the main pipeline function
for the abundance module `ppl_run_pipe_abu1()` is executed. This means that if we are
only interested in one species we can go ahead and use the `ppl_run_pipe_abu1()` function, directly.

To run the abundance module of the pipeline, it is enough to just run (`source()`) the script.
However, the default values will run the full pipeline for all species, and for
several years, which might take quite long. We will now go through the different
steps of the pipeline to better understand how to configure it to do what we want.

Note that while this is the default script we use for running the pipeline, there
is nothing special about it, and we could use something different that suits our needs.
The functionality of the pipeline comes from its functions.


### Configuration

```{r, eval=FALSE}

# Configure pipeline
config <- configPipeline(
    year = 2021,
    dur = 29,
    mod_file = "cwac_ssm_two_season_mean_rev_jump.R",
    package = "jagsUI",
    data_dir = "analysis/data",     # this might have to be adapted?
    out_dir = "analysis/output",    # this might have to be adapted?
    server = FALSE
)

# Read in catchment data. This should go as an argument
catchment <- sf::read_sf(file.path(config$data_dir, "quinary_catchmt_22.shp"))

# Re-project and simplify
catchment <- catchment %>%
    dplyr::select(QUATERNARY, Province, UNIT_ID) %>%
    sf::st_simplify(preserveTopology = TRUE, dTolerance = 1000) %>%
    sf::st_transform(crs = sf::st_crs(4326))

```

In the configuration section above we will create a `config` object using the `configPipeline()`
function that will be used throughout the pipeline by several functions.

The `configPipeline()` function, allows us to let the pipeline know what models
we want to run, for what years, what packages we want to use, etc. For detailed
information see `?configPipeline`.

Rather than specifying the covariates that we will use in our models, like we did
for the distribution module, here we only pass the model file name on to the function.
`configPipeline()` will look for the model file in `analysis/models`.

The only supported package at the moment is `jagsUI` and all package related functions
written for BIRDIE are stored in the `R/utils-jags.R` file.

(Note: there might be a better way of doing this, such as creating environment variables
or something like this, but this works for now)

The abundance module allow us to incorporate environmental covariates into the models,
although we are not using any at the moment. When we use covariates, we do so at the 
quinary catchment level. So rather than taking the value of the covariates at some specific
location we take the average value observed across the quinary catchment at some time.
We will see more of this on the *BIRDIE ABU: Data preparation* vignette. The important
point here is that we need to load the quinary catchment spatial object here to pass it on to the 
main pipeline function, so that it can be used during data preparation (edit the path
to the file).

### Create logs


```{r, eval=FALSE}

# Create log?
createLog(config, log_file = NULL, date_time = NULL, species = NA, model = NA,
          year = NA, data = NA, fit = NA, diagnose = NA, summary = NA,
          package = NA, notes = "Log file created")
```

The pipeline has a system to log all its activity. This is useful to keep track
of what species and years the pipeline has run for and whether there have been
any problems (e.g., species with too few data points or model fit errors). All the
activity is stored in .csv files that are saved to `analysis/output/reports`.

For more information check the logging functions of the BIRDIE package that are
stored on the `utils.R` file, notably see `?createLog()` for general logs and
`?logFitStatus()` for logs of model runs.

We see that for if we run the pipeline for several species, it makes sense to create
a log file only for the first species, and then use this file to store information
for all other species as well, each one in a row of the .csv file. Past this
first setup phase, the pipeline will look for the most recent log file and add information
to it, regardless of whether information is already present or not.


### Run modules


```{r, eval=FALSE}

for(i in 1:length(config$species)){

    sp_code <- config$species[i]

    message(paste0("Working on species ", sp_code, " (", i, " of ", length(config$species), ")"))

    # Run abudance pipeline module 1
    status_abu1 <- ppl_run_pipe_abu1(sp_code, config,
                                     steps = c("fit", "diagnose", "summary"),
                                     prep_data_steps = c("subset", "missing", "gee", "model"),
                                     summary_scale = "model",
                                     catchment = catchment,
                                     force_gee_upload = FALSE,
                                     force_gee = FALSE,
                                     monitor = TRUE)

    message(paste("ABU1 status =", status_abu1))

}

```

The final part of the script runs the main function of the abundance module of the pipeline,
looping through all species. The `ppl_run_pipe_abu1()` can run all the steps of the module: data preparation, model fitting, model diagnostics and model summary, or it can run just some of them. It will use the `config` object to know what package it should use for model fitting and the paths to
store model-ready data and model outputs. Note that some of the steps, may take quite long.
For example, to prepare data the pipeline connects to Google Earth Engine and annotates
data with environmental covariates, which can take a while. It also requires an internet
connection. Keep this in mind and note that we can skip some of the steps using the right
arguments. For more information see `?ppl_run_pipe_abu1`.
