---
title: "BIRDIE: Species distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE: Species distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```

## Introduction

There are two main analytical modules in BIRDIE: the distribution and abundance
modules.

In this document we will see how to use the distribution (DST) module. To do this we will 
have a look at the control script used to run the module and then we will break down
its different sections to understand what all the functions involved in this analysis do.

## The control script

The control script can be found at `/analysis/scripts/pipeline_script_dst.R`. This script
has three parts:

- Configuration
- Create logs
- Run modules

We will go through each of these parts below.

```{r, echo=FALSE, eval=FALSE}
writeLines(readLines("../analysis/scripts/pipeline_script_dst.R"))
```

The script is really just a for loop over species in which the main pipeline function
for the distribution module `ppl_run_pipe_dst1()` is executed. This means that we are
only interested in one species we can go ahead and use the `ppl_run_pipe_dst1()` function.

To run the distribution module of the pipeline, it is enough to just run the script.
However, the default values will run the full pipeline for all species, and for
several years, which might take quite long. We will now go through the different
steps the pipeline goes through to better understand how to configure it to do
what we want.

Note that while this is the default script we use for running the pipeline, there
is nothing special about it, and we could use something different that suits our needs.

### Configuration

```{r, eval=FALSE}

# We are currently working with several detection models
det_mods <- list(det_mod1 = c("(1|obs_id)", "log_hours", "prcp", "tdiff", "cwac"),
                 det_mod2 = c("(1|obs_id)", "(1|site_id)", "log_hours", "prcp", "tdiff", "cwac"))

# Configure pipeline
config <- configPipeline(year = 2010,
                         dur = 3,
                         occ_mod = c("log_dist_coast", "elev", "log_hum.km2", "wetcon",
                                     "watrec", "watext", "log_watext", "watext:watrec",
                                     "ndvi", "prcp", "tdiff"),
                         det_mod = det_mods$det_mod1,
                         fixed_vars = c("Pentad", "lon", "lat", "watocc_ever", "wetext_2018","wetcon_2018",
                                        "dist_coast", "elev"),
                         package = "spOccupancy",
                         data_dir = "analysis/data",     # this might have to be adapted?
                         out_dir = "analysis/output",    # this might have to be adapted?
                         server = FALSE)
```


In the piece of code above, we list a couple of possible detection models that we
can choose from (handy if we are planning of conducting several runs of the pipeline
with different models), and then we create a `config` object that will be used
throughout the pipeline by several functions.

The `configPipeline()` function, allows us to let the pipeline know what models
we want to run, for what years, what packages we want to use, etc. For detailed
information see `?configPipeline`.

(Note: there might be a better way of doing this, such as creating environment variables
or something like this, but this works for now)


### Create logs


```{r, eval=FALSE}

# Create log?
    if(i == 1){
        createLog(config, log_file = NULL, date_time = NULL, species = NA, model = NA,
                  year = NA, data = NA, fit = NA, diagnose = NA, summary = NA,
                  package = NA, notes = "Log file created")
    }
```

The pipeline has a system to log all its activity. This is useful to keep track
of what species and years the pipeline has run for and whether there have been
any problems (e.g., species with too few data points or model fit errors). All the
activity is stored in .csv files that are saved to `analysis/output/reports`.

For more information check the logging functions of the BIRDIE package that are
stored on the `utils.R` file, notably see `?createLog()` for general logs and
`?logFitStatus()` for logs of model runs.

We see that for if we run the pipeline for several species, it makes sense to create
a log file only for the first species (that is why `if(i == 1)`), and then use this file to store information
for all species, each in one row of the .csv file. If we just created one log for
each species, then there would be one .csv log file for each species. The past this
first setup phase, the pipeline will look for the most recent log file and add information
to it, regardless of whether information is already present or not.


### Run modules


```{r, eval=FALSE}

for(t in seq_along(config$years)){
    
    year_sel <- config$years[t]
    
    out_dst1 <- ppl_run_pipe_dst1(sp_code = sp_code,
                                  year = year_sel,
                                  config = config,
                                  steps = c("data", "fit", "summary", "diagnose"),
                                  force_gee_dwld = FALSE,
                                  monitor_gee = TRUE,
                                  force_site_visit = TRUE,
                                  force_abap_dwld = FALSE,
                                  spatial = FALSE,
                                  print_fitting = TRUE)
    
    message(paste("Pipeline DST1 status =", out_dst1))
    
    if(out_dst1 != 0){
        next
    }
    
}

```

The final part of the script runs the modules of the pipeline that we are interested in
(currently there is only one, because "module 2 - Area of Occupancy and other derived
indicators", is now run by the data base).
It will loop through the different years that we have set during the configuration
phase and it will run the main function of the distribution module `ppl_run_pipe_dst1()`.
This function can run all the steps of the module: data preparation, model fitting, 
model diagnostics and model summary, or it can run just some of them. It will use the
`config` object to know what package it should use for model fitting and the paths to
store model-ready data and model outputs. Note that some of the steps, may take quite long.
For example, to prepare data the pipeline connects to Google Earth Engine and annotates
data with environmental covariates, which can take a while. It also requires an internet
connection. Keep this in mind and note that we can skip some of the steps using the right
arguments. For more information see `?ppl_run_pipe_dst1`.
