---
title: "BIRDIE ABU: Data preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE ABU: Data preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BIRDIE)
```

## Introduction

The species abundance module (ABU) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species abundance* vignettes for general details
about BIRDIE and about the ABU module, respectively. In this vignette, we will go through
the different tasks that are performed during the first step of the ABU module: **data preparation**.

The main function used for data preparation is `ppl_create_data_ssm()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work.

Data preparation has three main tasks: 

- Download CWAC data and subset sites suitable for modelling
- Complete the dataset by adding missing counts (with NA)
- Annotate count data with environmental covariates
- Format data for state-space modelling

By just running `ppl_create_data_ssm()` all of the tasks will be performed and our
data will be prepared. However, to understand what goes on "under the hood", we will
explain each of these tasks, and how they are conducted, below.

Note that data preparation can be time-consuming and it is more efficient to prepare
data for multiple years at once. We are currently fitting 20 years of data at a time,
and so that is the number of years that will be prepared by `ppl_create_data_ssm()`. The number of 
years prepared is given by the `dur` argument of the `configPipeline()` function.


## Download CWAC data and subset sites

There is a first section in the `ppl_create_ssm_data()` function that run when we tell
the function to subset sites or to complete the dataset with missing counts. In this
section we will download CWAC data using the function `CWAC::getCwacSppCounts()` from 
the [`CWAC` R package](https://github.com/AfricaBirdData/CWAC). We then proceed to
include any data that we have and are not on the CWAC data base. At the moment of
writing we have data from DuToit's pan contributed by Doug Harebottle. These data
is formatted and incorporated to the data downloaded from CWAC. If we had any other
data we wanted to include we would need to modify this part of the `ppl_create_data_ssm()`
function code.

Once all of the data is combined we subset those sites that have been counted at least
five times during summer and five times during winter between the years 1993 and 2021.
Those species that don't meet the requirements should be analysed differently, although
we still don't have an alternative model for them. During model diagnosis there is
another filter where model outputs with too large of a difference between the estimates
and the upper limit of the credible intervals are also discarded (see BIRDIE ABU: Model diagnostics).

## Adding missing counts

When data comes out of the CWAC database there is no reference to missing counts, meaning
that if in any year nobody went to count a certain wetland during a certain season this data point
would just be absent from the data set. What we would like instead is a record for that season
and year with a missing (`NA`) count. This is convenient for multiple reasons, but perhaps
the most important one is that `JAGS` will automatically treat these missing data points
as parameters that need to be estimated. We give missing summer counts a date that 
corresponds to the first day of January (perhaps we should reconsider this, because
summer counts only start on the 15th of January) and to winter counts we assign the first
of July.

## Annotate with environmental covariates

Although we are not currently using covariates in our modelling, we may use environmental
covariates to model abundance, which requires count data to be annotated with this information. To 
facilitate the automation of this process and periodic updates when new data becomes
available, we use the data sets and functionality offered by Google Earth Engine (GEE).

The functionality to connect and transfer data to/from GEE is provided by the [`ABDtools` R package](https://github.com/AfricaBirdData/ABDtools).
This package basically wraps functions from [`rgee`](https://github.com/r-spatial/rgee); another package on which
it depends heavily. Therefore, it is a requirement to have `rgee` properly installed and configured
to be able to perform data-annotation tasks. Check the GitHub repos for [`rgee`](https://github.com/r-spatial/rgee)
and [`ABDtools`](https://github.com/AfricaBirdData/ABDtools).

Once these two packages are installed and configured, we can use their functionality in the pipeline.
In BIRDIE we use the function `prepGEECatchmData()` to annotate CWAC data. See `?prepGEECatchmData()` for details.
The function makes reference to "catchment" because at the moment this function is
prepared to use the quinary catchment CWAC sites are located at as a reference area for the covariates.
So rather than extracting environmental information from some specific point location, we extract
all pixels contained in the quinary catchment and we take the average value of the 
covariate across those pixels.

Annotating with different variables using GEE requires different procedures. Therefore,
there is no way to flexibly communicate to these functions which variables we want to
annotate our data with. Instead, we have **hard-coded** the variables we are using for the
BIRDIE pipeline. If we wanted to change the variables we use, then we would have to
modify the `prepGEECatchmData()` function. This is not ideal,
but it is how it is set up currently.

One important thing to keep in mind is that we consider that waterbird summer populations
should be affected by the environmental conditions of the previous year, rather than 
on the same year. This is because summer occur in January and therefore the average
conditions on the previous year are likely to affect summer populations more directly
than those on the same year, which have still not presented themselves at the time of
counting.

Another important thing to keep in mind is that some environmental layers used don't have
information past a certain date. We have set up the functions in such a way that data past
the last date of the layer get annotated with the latest available information (last date of the layer).
Whenever the pipeline is run it is advised to review the environmental layer used and 
the last date information is available for, and update the functions if necessary.

## Format data for state-space modelling

In a final step we prepare the data for modelling. We will not format it to fit into
any specific package yet. Here we create certain variables that are useful such
as ids for site, year and visit.

Very importantly, here there is a choice to make in terms of what to do when seasonal
counts are duplicated. For now, we keep all counts that are labelled as `summer` or 
`winter` counts in the CWAC data and consider them to be replicates.
