---
title: "BIRDIE ABU: Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BIRDIE ABU: Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BIRDIE)
```

## Introduction

The species abundance module (ABU) of the BIRDIE pipeline has four main steps:
data preparation, model fitting, model diagnostics and model summary. See the 
*BIRDIE: basics* and *BIRDIE: species abundance* vignettes for general details
about BIRDIE and about the DST module, respectively. In this vignette, we will go through
the different tasks that are performed during the step of the DST module: **model diagnostics**.

The main function used for model diagnosis is `ppl_diagnose_ssm()`. This is a
`ppl_` function, and therefore it doesn't do much processing itself (see *BIRDIE: basics* if
this is confusing), but it does call the right functions to do the work.

During the diagnostics step we look at two indicators: the Gelman-Rubin convergence
diagnostic ($\hat{R}$) and a goodness-of-fit Bayesian p-value. The first diagnostic is
computed by the function `diagnoseRhatJagsSsm()` and the second one by the function
`diagnoseGofJagsSsm()` (both functions are found in `R/utils-jags.R`).

`ppl_diagnose_ssm()` combines the output of these two functions into a single table
that is stored in `analysis/output/sp_code/abu_diagnostics_spOccupancy_YYYY_sp_code.csv`,
where `YYYY` is the year of the data we are fitting a model to and `sp_code` is the
species code. In this way, it is easy to extract all diagnostic files from the different
directories, combine them into a table and quickly explore potential issues.

## Diagnosing convergence

The function `diagnoseRhatJagsSsm()` extracts $\hat{R}$ values computed by `JAGS`
for each of the estimated parameters and tabulates them. In this table, each parameter
appears in one column and gets a 0 if $|1 - \hat{R}| < 0.1$ and a 1 otherwise. Some
extra columns are added, such as number of non-convergent parameters, number of detections
observed, number of observations, etc.

## Diagnosing goodness-of-fit

The function `diagnoseGofJagsSsm()` conducts posterior predictive checks (PPC) and calculates
a Bayesian p-value.

The procedure consists of simulating a number of count data sets
from the posterior distribution and comparing them with
the observed data. 

We calculate the mean and standard deviation for each of the simulated data sets. Then,
we calculate the number of data sets that are below and above the mean observed in the
data ("Tmean"). If we obtain proportions close to 0.5 it means that our predictions are unbiased.
We follow a similar procedure with the standard deviations ("Tsd") to see if our predictions
have a similar dispersion as that observed in the data. Finally, for each simulation we
calculate the mean difference observed with the data ("Tdiff"), which should give us an idea of the
magnitude of the bias, should it exist.

## Diagnosing uncertainty

Streamlined analyses can be difficult to stabilize for all species and sites. Sometimes
state estimate uncertainty can grow quickly, particularly in long periods with
missing data at the beginning and end of the time series. We are particularly 
concerned with uncertainty exploding towards recent years, because this can impact
managers and decision makers. Therefore, we test whether the upper limit of the 
confidence band estimated for the last 10 years ever gets higher than
50 times the highest estimate for that period, at any site for any species. If this
is the case we exclude estimates and uncertainty bands for those sites and species
combinations. We still present counts.

The function `ppl_diagnose_ssm()` identifies for each species what sites present
large uncertainty at the end of the time series and stores them in the diagnostics
file. This information will then be used by the summary function to exclude the 
corresponding estimates.

## Shiny apps

We have developed a Shiny app that allows us to extract diagnostics for any species
and year and explore them in an interactive way. The app is still experimental and
can be found at `analysis/tools/diagnose_dabu`
